`r opts_chunk$set(cache=TRUE, fig.width=10)`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

# Example Data for phyloseq

```{r, message=FALSE, warning=FALSE}
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")
```

ggplot2 package theme set. See [the ggplot2 online documentation](http://docs.ggplot2.org/current/) for further help.

```{r ggplot2-theme-set}
theme_set(theme_bw())
```


# Example Data
## Included Example Data
There are multiple example data sets included in phyloseq. Many are from published investigations and include documentation with a summary and references, as well as some example code representing some aspect of analysis available in phyloseq.

To load example data into the working environment, use the `data()` command:

```{r eval=FALSE}
data(GlobalPatterns)
data(esophagus)
data(enterotype)
data(soilrep)
```

## Example Data Documentation
In the package index, go to the names beginning with "data-" to see the documentation of currently available example datasets.

Try for example

```{r eval=FALSE}
?GlobalPatterns
```

to access the documentation for the so-called "GlobalPatterns" dataset.

## Example Script Using Example Data
You can also try the examples included with the example data documentation (as well as examples for functions/methods) using the standard `example` command in R -- in this case the examples for the `enterotype` dataset.

```{r run-examples}
example(enterotype, ask=FALSE)
```



---

### Other tutorial pages for the phyloseq package:

```{r other-graphics-links, results='asis', echo=FALSE}
htmlfiles = list.files(pattern="\\.html")
htmlstems = gsub(".html", "", htmlfiles)
mdlines = paste("#### [", htmlstems, "](", htmlfiles, ")", sep="")
cat(mdlines, sep="\n\n")
```
