`r opts_chunk$set(cache=TRUE, fig.width=8)`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>


Subset points in an ordination plot
========================================================

The `subset_ord_plot` function is a "convenience function" intended to make it easier to retrieve a plot-derived `data.frame` with a subset of points according to a `threshold` and `method`. The meaning of the `threshold` depends upon the `method`.

Load the phyloseq package, and the GlobalPatterns example dataset

```{r, message=FALSE}
library("phyloseq")
data("GlobalPatterns")
```
For completeness, here is the version number of phyloseq used to build this instance of the tutorial – and also how you can check your own current version from the command line.

```{r}
packageDescription("phyloseq")$Version
```

Some subsetting and light massaging of the `GlobalPatterns` dataset.
```{r}
# Need to clean the zeros from GlobalPatterns:
GP <- GlobalPatterns
GP <- prune_species(taxa_sums(GP)>0, GP)
# Add "human" variable to GP
sample_data(GP)$human <- get_variable(GP, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")
# Subset to just Bacteroidetes
GP <- subset_taxa(GP, Phylum=="Bacteroidetes")
```

Perform a correspondence analysis and then create some plots to demonstrate using `subset_ord_plot`. We want to make species topo with a subset of points layered. Start by performing the correspondence analysis.

```{r}
gpca <- ordinate(GP, "CCA")
```

Now make a basic plot of just the species points in the correspondence analysis.
```{r}
p1 = plot_ordination(GP, gpca, "species", color="Class")
p1
```

Re-draw this as topo without points, and facet
```{r}
library("ggplot2")
p0 = ggplot(p1$data, p1$mapping) + geom_density2d() + facet_wrap(~Class)
p0
```

Re-draw this but include points
```{r}
p1 = p1 + geom_density2d() + facet_wrap(~Class)
p1
```

Add a layer of a subset of species-points that are furthest from origin.
```{r}
p0 + geom_point(data=subset_ord_plot(p1, 0.7, "square"), size=1)
p0 + geom_point(data=subset_ord_plot(p1, 0.7, "farthest"), size=1)
p0 + geom_point(data=subset_ord_plot(p1, 0.7, "radial"), size=1)
```

Here is what the data retreived by `subset_ord_plot` actually looks like
```{r}
head(subset_ord_plot(p1, 0.7, "radial"))
```

